A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-11 calculus.tex
moved basic automata definitions to section 3

We start by defining our calculus, \Calculus{}, for one automaton, and only
extend it to products of automata in \cref{sec:multiple}. We assume basic
familiarity with first-order logic, Presburger arithmetic, and the classical
sequent calculus. For reference, see e.g. \cite{Fitting96a}.

First, assume an NFA $\Automaton = \AutomatonTuple$ with $\NrTransitions$
transitions $\Transitions =
\Set{\EllipsisSequence{\Transition}{\NrTransitions}}$. For convenience, we then
introduce the following supporting notations:

\begin{definition}
  A \textit{path} $\Path = \PathEnumeration$ of an automaton~$\Automaton$ with
  transitions $\Transitions$ represents a path through $\Automaton$ using
  transitions in $\Transitions$ (i.e. $\Transitions(\State_k, \Label_{k+1},
  \State_{k+1})$ holds), passing zero or more labels $\Label_1, \ldots,
  \Label_n$. The path must begin in the initial state, i.e.~$\State_0 =
  \InitialState$. The end state, $\State_n$, is not necessarily accepting. If it
  is, we say the path is also \textit{accepting}.
  \end{definition}

\begin{definition}
  Moreover, we talk about the \textit{set of paths} of an automaton,
  $\Paths(\Automaton)$, a possibly infinite (if $\Automaton$ has loops) set of
  valid paths through $\Automaton$. Additionally, we use the
  notation~$\Paths(\Automaton, \State)$ to mean all paths ending in
  state~$\State$.
\end{definition}

\begin{definition}
  For a path $\Path = \Tuple{\State_0 \Label_1 \State_1 \RepeatSum{,}
  \State_n}$, its \textit{word} $\WordOf(\Path) = \Label_1 \RepeatSum{\Concat}
  \Label_n$ is the word read out on its labels.
\end{definition}

\begin{definition}
  The \textit{states} of a path $\Path = \Tuple{\State_0 \Label_1 \State_1
  \RepeatSum{,} \State_n}$, $\StatesOf(\Path) = \Set{\State_0\RepeatSum{,}
  \State_n}$ are the states visited along $\Path$. Note that $\InitialState \in
  \StatesOf(\Path)$ for every path since all paths start in the initial state.
\end{definition}

\begin{definition}
  A \textit{dominating set of transitions}, $C$ of an automaton~$\Automaton$ of
  a transition $\Transition$, written $\Dominates(C, \Automaton, \Transition)$,
  is a set of transitions from $\Automaton$ such that every accepting path $p
  \in \Automaton$ where $\Transition \in p$ contains at least one transition
  from $C$. Notably, $\Dominates(\emptyset, \Automaton, \Transition')$ for every
  dead transition $\Transition'$ and $\Dominates(Set{\Transition''}, \Automaton,
  \Transition)$.
\end{definition}

\begin{definition}
 The \textit{transition count}, $\TransitionCount(\Transition, \Path)$ is the
 number of times a transition $\Transition =
 \FromLabelTo{\State_1}{\Label}{\State_2} \in \Transitions$ appears in a path
 $p$.
\end{definition}

@@

\begin{definition}\label{def:single-image}
  The Parikh predicate, $\SinglePredicateInstance$, for some automaton
  $\Automaton = \AutomatonTuple$, modulo some map $\Map : \Alphabet^* \to \Monoid$ to a
  commutative monoid $\Monoid$ as described in \cref{sec:generalised}
  and with a transition selection function
  $\Filter:\: \Transitions \to \Naturals$ holds when
  $\MonoidElement \in \Monoid$ is an element of the Parikh image of
  $\Language(\Automaton)$
  modulo $\Map$, or more formally when there is an accepting path
  $\Path = \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ such
  that $\Filter(\Transition) = \TransitionCount(\Transition, \Path)$
  for all $t \in \Transitions$, and $\MonoidElement = \Map(\WordOf(\Path))$.
\end{definition}

\begin{definition}\label{def:connected} $\Connected(\Automaton, \Filter)$ for
  some automaton $\Automaton = \AutomatonTuple$ holds when for every
  $\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions$,
  $\Filter(\Transition) > 0 \implies \exists \Path \in \Paths(\Automaton)$
  $\Filter(\Transition_i) > 0$ for $\Transition_i \in p$, and $\State \in
  \StatesOf(p)$, or in words that there exists some $\Filter$-selected valid
  path that reaches $\Transition$'s starting state, $\State$. It represents the
  condition that $\Automaton$ is connected with respect to the selection
  function $\Filter$ for every transition. It is redundant to $\Image$ by
  design.
\end{definition}

>>>>

We start by defining our calculus, \Calculus{}, for one automaton, and only
extend it to products of automata in \cref{sec:multiple}.
%
Assume an automaton $\Automaton = \AutomatonTuple$ with $\NrTransitions$
transitions $\Transitions =
\Set{\EllipsisSequence{\Transition}{\NrTransitions}}$. We use
the notations introduced in \cref{sec:languages}. For convenience, we
introduce the following additional notions:

\begin{definition}
  A set~$C$ of transitions of an automaton~$\Automaton$ \emph{dominates}
  a  transition~$\Transition$, written
  $\Dominates(C, \Automaton, \Transition)$, if
  every accepting path $p$ of $\Automaton$
  with $\Transition \in p$ contains at least one transition from
  $C$. Notably, $\Dominates(\emptyset, \Automaton, \Transition)$ for
  every unreachable transition $\Transition$, and
  $\Dominates(\Set{\Transition}, \Automaton, \Transition)$ for every
  transition~$\Transition$.
\end{definition}

\begin{definition}
  The \textit{transition count},
  $\TransitionCount(\Transition, \Path)$ is the number of times a
  transition
  $\Transition = \FromLabelTo{\State}{\Label}{\State'} \in
  \Transitions$ appears on a path $p$. A \emph{transition selection
    function} is a function~$\Filter:\: \Transitions \to \Naturals$
  labelling every transition~$\Transition \in \Transitions$ with a
  non-negative number.
\end{definition}

@@

\begin{definition}\label{def:single-image}
  The Parikh predicate, $\SinglePredicateInstance$ holds for some
  automaton $\Automaton = \AutomatonTuple$, some map
  $\Map : \Alphabet^* \to \Monoid$ to a commutative monoid $\Monoid$,
  some transition selection function
  $\Filter:\: \Transitions \to \Naturals$, and some monoid
  element~$\MonoidElement \in \Monoid$ if $\MonoidElement$ is an
  element of the Parikh image of $\Language(\Automaton)$ modulo
  $\Map$, or more formally when there is an accepting path
  $\Path = \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ such
  that $\Filter(\Transition) = \TransitionCount(\Transition, \Path)$
  for all $t \in \Transitions$, and
  $\MonoidElement = \Map(\WordOf(\Path))$.
\end{definition}

\begin{definition}\label{def:connected} $\Connected(\Automaton, \Filter)$
  holds for some automaton $\Automaton = \AutomatonTuple$ and
  transition selection function
  $\Filter:\: \Transitions \to \Naturals$
  if for every
  $\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions$,
  if $\Filter(\Transition) > 0$ then there is some $\Filter$-selected
  path that reaches $\Transition$'s starting state~$\State$. More formally,
  if $\Filter(\Transition) > 0$ then there is some
  path~$\Path \in \Paths(\Automaton)$ with
  $\Filter(\Transition') > 0$ for every $\Transition' \in p$, and $\State \in
  \StatesOf(p)$. Predicate~$\Connected(\Automaton, \Filter)$ represents the
  condition that $\Automaton$ is connected with respect to the selection
  function $\Filter$ for every transition, and is implied by
  $\SinglePredicateInstance$.
\end{definition}

### 2023-07-11 preliminaries.tex
more
moved basic automata definitions to section 3

@@

\subsection{Languages, Finite-state Automata and their Products}

@@

A \emph{finite-state automaton}~$\Automaton$ with alphabet~$\Alphabet$ is a
tuple~$\AutomatonTuple$, where $\States$ is the set of states,
$\InitialState$ the initial state, $\AcceptingStates$ the set of
accepting states, and
$\Transitions \subseteq \States \times \Alphabet \times \States$ the
transition relation.  We write a transition
$\Transition = \Tuple{\State, \Label, \State'} \in \Transitions$ as
$\Transition = \FromLabelTo{\State}{\Label}{\State'}$.  Similarly, we
use the notation $\FromLabelTo{\State}{}{}$ to refer to the set of
transitions starting in $\State$, and $\FromLabelTo{}{}{\State}$ to
refer to the set of transitions coming into $\State$, whenever the
automaton is clear from the context. An accepting run of $\Automaton$
is a sequence
$\InitialState = q_0 \xrightarrow{l_1} q_1 \xrightarrow{l_2} \cdots
\xrightarrow{l_k} q_k \in \AcceptingStates$, and the accepted word is
$l_1 l_2 \ldots l_k \in \Strings$. The set of words accepted by an
automaton~$\Automaton$ is denoted by
$\Language(\Automaton) \subseteq \Strings$.

@@

>>>>

We first survey some of the required background on finite-state
automata and the Parikh image. In addition, we assume basic familiarity
with first-order logic, Presburger
arithmetic, and the classical sequent calculus; for reference, see
e.g. \cite{Fitting96a}.

@@

\subsection{Languages, Finite-state Automata and their Products}
\label{sec:languages}

@@

A \emph{finite-state automaton}~$\Automaton$ with alphabet~$\Alphabet$ is a
tuple~$\AutomatonTuple$, where $\States$ is the set of states,
$\InitialState$ the initial state, $\AcceptingStates$ the set of
accepting states, and
$\Transitions \subseteq \States \times \Alphabet \times \States$ the
transition relation.  We write a transition
$\Transition = \Tuple{\State, \Label, \State'} \in \Transitions$ as
$\Transition = \FromLabelTo{\State}{\Label}{\State'}$.  Similarly, we
use the notation $\FromLabelTo{\State}{}{}$ to refer to the set of
transitions starting in $\State$, and $\FromLabelTo{}{}{\State}$ to
refer to the set of transitions coming into $\State$, whenever the
automaton is clear from the context.

@@

A \textit{path} $\Path = \PathEnumeration$ of an
automaton~$\Automaton$ is a sequence of
states~$\State_0, \ldots, \State_n$ interleaved with
letters~$\Label_1, \ldots, \Label_n$ such that
$\State_0 = \InitialState$ and
$\FromLabelTo{\State_{i-1}}{\Label_{i}}{\State_{i}}$ for
$i \in \{1, \ldots, n\}$.  The path is \emph{accepting} if the end
state is accepting, $\State_n \in \AcceptingStates$. The \textit{set
  of paths} of $\Automaton$ is denoted by
$\Paths(\Automaton)$. Additionally, we use the
notation~$\Paths(\Automaton, \State)$ to mean all paths ending in
state~$\State$. We write $\Transition \in \Path$ to express that
transition~$\Transition$ occurs on path~$\Path$. The \textit{states} of a path
$\Path = \PathEnumeration$, denoted
$\StatesOf(\Path) = \Set{\State_0\RepeatSum{,} \State_n}$ are the
states visited along $\Path$. Note that
$\InitialState \in \StatesOf(\Path)$ for every path since all paths
start in the initial state.  The \textit{word} of a path
$\Path = \PathEnumeration$ is the word
$\WordOf(\Path) = \Label_1 \RepeatSum{\Concat}
 \Label_n \in \Strings$ formed by the labels on the path. Finally, the
 set of words accepted by an automaton~$\Automaton$, denoted by
 $\Language(\Automaton) \subseteq \Strings$, is the set of words of
 accepting paths.
